NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-02-06 Manuscript/manuscriptPS.tex
Update simulation details
Stylistics tuning of discussion.

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of cation binding
are crucial for the PS mediated signaling processes.

Previous experimental studies have concluded that
PS headgroups are more rigid than phophocholines (PC)
due to the hydrogen bonding network or
electrostatic interactions \cite{browning80,buldt81}.
Multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids,
while monovalent ions interact more weakly with PS
containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
The dehydrated complexes of PS headgroup and calcium ions can also lead to the
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
On the other hand, some studies propose that the specific binding
affinity is similar to the negatively charged and zwitterionic lipids and that
the increased cation binding to negatively charged lipid bilayer arise only due
to the increase of local cation concentration in the vicinity of membranes \cite{seelig90,sinn06}.
Dilution of bilayers with PC lipids makes PS headgroups
less rigid and reduces propensity for the formation of
strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.
The molecular level interpretation of these observations is,
however, not available.

@@

Several classical molecular dynamics (MD) simulation studies are done
to understand PS headgroups, their influence on lipid bilayer properties and
interactions with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
However, the results strongly depend on the used force field parameters.
For example, the recent simulations using NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13} combined with 2D infrared spectroscopy
suggests that calcium ions interacts only with the carboxylate group of PS lipids \cite{valentine18}, while
the same force field without the NBfix parameters together with the NMR chemical shifts and
REDOR experiments suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04}
combined with fluorescent and vibrational sum frequency spectroscopy suggested a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}. 
We have recently demonstrated that such controversies can be resolved by
comparing C-H bond order parameters of lipid headgroups between simulations
and experiments \cite{botan15,catte16}, which can be directly
measured from NMR experiments with high accuracy and compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}.
Previous studies showed that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most simulation models \cite{botan15} and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}. Based on this data, the cation binding
affinity to POPC bilayer was then improved by implicitly including the electronic polarizability
using the electronic continuum correction \cite{melcr18}.

Here, we collect the set of experimentally measured lipid headgroup and
glycerol backbone C-H bond order parameters, which can be used to
evaluate the quality of headgroup structure and the ion binding affinity
in MD simulations of lipid bilayers containing PS lipids. 
The available MD simulation models of PS are then compared with
the collected experimental data using the NMRlipids open collaboration
project (\url{www.nmrlipids.blogspot.fi}). The results pave the way
for the development of lipid force fields with realistic description of
the headgroup region of negatively charged lipids in physiological salt
conditions. Such models are expected to be useful in understanding
biological function of lipid headgroups and glycerol backbone because
they behave similarly in model membranes and in bacterial cells \cite{gally81,scherer87,seelig90}.

@@

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
    The order parameter values from systems without calcium are set as the zero point of y-axis,
    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
    such that the lower order parameter is zero without additional ions to correctly illustrate
    the forking with different concentrations of calcium.
    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
  }
  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056.
  these are not necessary, but could be added here if delivered.} \\
\end{figure*}

>>>>

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eukaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of red blood cells,
but the abundance varies between different organelles up to
25-35\% in the cytosolic leaflet of plasma membranes \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance \todo{Relatively low compared to what?}, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains interact specifically with PS lipids,
while others are attracted by nonspecific electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of cation binding
are crucial for the PS-mediated signaling processes.

Previous experimental studies have concluded that the
PS headgroup is more rigid than the phosphocholine (PC)
due to the hydrogen bonding network or
electrostatic interactions \cite{browning80,buldt81}.
Multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids,
while monovalent ions interact more weakly with
PS-containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
The dehydrated complexes of PS headgroup and calcium ions can also lead to
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
On the other hand, some studies propose that the specific binding
affinity is similar for the negatively charged and zwitterionic lipids, and
the increased cation binding to negatively charged lipid bilayers arises only due
to the increased local cation concentration in the membranes vicinity~\cite{seelig90,sinn06}.
Dilution of bilayers with PC lipids makes PS headgroups
less rigid and reduces their propensity to form
strong complexes with multivalent ions~\cite{browning80,buldt81,roux90,roux91}.
The molecular level interpretation of these observations is,
however, lacking.

@@

Several classical molecular dynamics (MD) simulation studies have been done
to understand PS headgroups, their influence on lipid bilayer properties, and their
interaction with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
However, the results depend strongly on the force field used.
For example, the recent simulations using NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13}, combined with 2D infrared spectroscopy,
suggest that calcium ions interact only with the carboxylate group of PS lipids \cite{valentine18}; in contrast,
the same force field without the NBfix parameters, combined with NMR chemical shifts and
REDOR \todo{Spell out REDOR?} experiments, suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04},
combined with fluorescent and vibrational sum frequency spectroscopy, suggest a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.
We have recently demonstrated that such controversies can be resolved by
comparing the C--H bond order parameters, $S_\mathrm{CH}$, of lipid headgroups between simulations
and experiments \cite{botan15,catte16}. The $S_\mathrm{CH}$ can be directly
measured from NMR experiments with high accuracy and compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}.
Previous studies showed that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most MD force fields~\cite{botan15}, and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}. Based on these data, the cation binding
affinity to POPC bilayer has since been improved by implicitly including the electronic polarizability
using the electronic continuum correction \cite{melcr18}.

Here, we collect a set of experimentally measured lipid headgroup and
glycerol backbone C-H bond order parameters, which can be used to
evaluate the quality of headgroup structure and the ion binding affinity
in MD simulations of lipid bilayers containing PS lipids. 
The available MD simulation models of PS are then compared with
the collected experimental data.
%using the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
The results pave the way
for the development of lipid force fields with realistic description of
the headgroup region of negatively charged lipids in physiological salt
conditions. Such models are expected to be useful in understanding
biological function of lipid headgroups and glycerol backbone because
they behave similarly in model membranes and in bacterial cells \cite{gally81,scherer87,seelig90}.

@@

    POPC:POPS (1:0)        & MacRog \cite{maciejewski14} &0    & 120:0  & 5120 & 0    & 298  & 200 & 150 & \cite{macrogPOPC298K}  \\

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration from experiments \cite{roux90} and different simulations
    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
    The order parameter values from systems without calcium are set as the zero point of y-axis,
    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
    such that the lower order parameter is zero without additional ions to correctly illustrate
    the forking with different concentrations of calcium.
    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
  }
%  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
%    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056.
%  these are not necessary, but could be added here if delivered.} \\
\end{figure*}
